Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) \note{\scriptsize Authors, please add/check middle initials and affiliations}

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
finishing conclusions for first draft

The papers submitted to this panel included many general
recommendations for processes, practices, tools, etc. One of the papers
suggested that a ``Software Sustainability Institute'' should be vested with
the authority to develop standardized quality processes, a central common
repository, central resources for services and consulting, a think tank of
sorts, and a software orphanage center. \note{What is software orphanage center?} The idea of one common repository received
some resistance, with so many compelling alternatives available, e.g. Bitbucket or
GitHub. The centralized communication or point of contact was seen as
reasonable, with the statement that ``vested with authority'' is perhaps too strong,
but ``providing tools if needed'' might be more appropriate.

@@

To conclude, we discuss the major concerns and issues that were raised in
the workshop, what we have learned, and what we plan to do going forward.

\note{add some analysis of the deficiencies and difficulties that are
  present in different fields, and those that are common?}

\subsection{Concerns and issues}

\todo{what are the main issues discussed in each discussion section? - and, if you want, and recommendations on what should be done about them}

The issues that came up repeatedly in different parts of the workshop are, as described in
\S\ref{sec:cross-cutting}, the need for a definition of sustainability, and the need for
career paths for scientific software developers.

\subsubsection*{Developing and Supporting Software}

\subsubsection*{Policy}

\subsubsection*{Communities, Models, and Education}

Recognition of scholarship in scientific software over research 
articles is overdue.

\subsection{Recommendations or lessons}

\note{lessons from the 3 discussion sessions go here, and lessons from the cross-cutting section as well?}

\subsubsection*{Developing and Supporting Software}

Successful collaborative projects conducted via the use of shared
repositories ensures reproducibility in computational science.

\subsubsection*{Policy}

One of the subjects that contributed papers, attendees, and speakers
wrestled with repeatedly is defining sustainability such that the
community can get behind it. Although some had hoped that at least an
initial consensus could be reached in the course of the workshop, this
proved elusive. However, in the absence of such a definition it will
remain difficult to define exactly what the goals should be towards
achieving, or even only improving software sustainability, and hence
what practices should be followed and implemented when. Thus, to be
seen successful, activities emerging from or following this first
WSSSPE workshop should include working out a community consensus on
what we mean by sustainability.

\subsubsection*{Communities, Models, and Education}
A sustainability model for scientific software is to build a pipeline
from construction to consumption as found in the most efficient
information technology enterprises.

% Conclusion of S8 : Use Cases, not sure where to put
\note{conclusions from S8, not sure of location --Ketan}

Two distinct class of scientific software projects and products were discussed
in the workshop: 1) generic, large-scale utility software and 2) niche, medium-
and small-scale scientific software. Each class faces different and significant
challenges. New undertakings should recognize the differences in advance and
identify such challenges within the development and sustaining efforts. In
particular, the dynamics associated with developers, scope, life cycle,
users-community, (re)usability, funding support and career paths vary widely
among the two classes of software.

@@

\subsection{Future activities}

>>>>

The papers submitted to this panel included many general
recommendations for processes, practices, tools, etc. One of the papers
suggested that a ``Software Sustainability Institute'' should be vested with
the authority to develop standardized quality processes, a central common
repository, central resources for services and consulting, a think tank of
sorts, and a software orphanage center (i.e., a place to `take care' of software
when the original developers have stopped doing so).
The idea of one common repository received
some resistance, with so many compelling alternatives available, e.g. Bitbucket or
GitHub. The centralized communication or point of contact was seen as
reasonable, with the statement that ``vested with authority'' is perhaps too strong,
but ``providing tools if needed'' might be more appropriate.

@@

To conclude, we discuss what we have learned from the workshop and what we plan to do going forward.

%\note{add some analysis of the deficiencies and difficulties that are
%  present in different fields, and those that are common?}

%\subsection{Concerns and issues}
%
%\todo{what are the main issues discussed in each discussion section? - and, if you want, and recommendations on what should be done about them}

\subsection{Issues and lessons}

%\note{lessons from the 3 discussion sessions go here, and lessons from the cross-cutting section as well?}

Three issues came up repeatedly in different parts of the workshop:

\begin{enumerate}
\item The need for a definition of sustainability such that the
community can get behind it. Although some had hoped that at least an
initial consensus could be reached in the course of the workshop, this
proved elusive. However, in the absence of such a definition it will
remain difficult to define exactly what the goals should be towards
achieving, or even only improving software sustainability, and hence
what practices should be followed and implemented when. As described
in the next subsection (\S\ref{sec:future}), the workshop organizers have begun an effort
to address this.

\item The need for academic
career paths for scientific software developers.  
Unfortunately, it is not clear how to ensure that these career paths become available,
other than repeatedly talking about this issue. The recent Moore and Sloan initiative
in data science~\cite{moore_sloan} are trying to address this, to some extent, by providing funds and
incentives to universities in the US that work towards this goal.

%
%\subsubsection*{Developing and Supporting Software}
%
%\subsubsection*{Policy}
%
%\subsubsection*{Communities, Models, and Education}

\item The need for recognition of scholarship in scientific software over research 
articles. This need probably is the most addressed of the three, today, with
efforts underway such as the Mozilla Science Lab, GitHub, and figshare
``Code as a research object'' project~\cite{code_as_a_research_object} among others.

\end{enumerate}

%\subsubsection*{Developing and Supporting Software}

\subsubsection*{Discussion sessions}

Two strong lessons came out of the three discussion sessions:

\begin{enumerate}
\item
Successful collaborative projects conducted via the use of shared
repositories are a means towards reproducibility in computational science.
\katznote{this seems rather limited - what else comes out of `successful collaborative projects conducted via the use of shared
repositories'?}

%\subsubsection*{Policy}

%One of the subjects that contributed papers, attendees, and speakers
%wrestled with repeatedly is defining sustainability such that the
%community can get behind it. Although some had hoped that at least an
%initial consensus could be reached in the course of the workshop, this
%proved elusive. However, in the absence of such a definition it will
%remain difficult to define exactly what the goals should be towards
%achieving, or even only improving software sustainability, and hence
%what practices should be followed and implemented when. As described
%in the next subsection, a subset of the organizers have begun an effort
%to address this.

%Thus, to be
%seen successful, activities emerging from or following this first
%WSSSPE workshop should include working out a community consensus on
%what we mean by sustainability.

%\subsubsection*{Communities, Models, and Education}
\item A sustainability model for scientific software is to build a pipeline
from construction to consumption, as found in the most efficient
information technology enterprises.

\end{enumerate}

% Conclusion of S8 : Use Cases, not sure where to put
%\note{conclusions from S8, not sure of location --Ketan}

\subsubsection*{Use cases}
Two distinct class of scientific software projects and products were discussed
in the workshop: 1) generic, large-scale utility software and 2) niche, medium-
and small-scale scientific software. Each class faces different and significant
challenges. New undertakings should recognize the differences in advance and
identify such challenges within the development and sustaining efforts. In
particular, the dynamics associated with developers, scope, life cycle,
users-community, (re)usability, funding support and career paths vary widely
among the two classes of software.

@@

\subsection{Future activities} \label{sec:future}
